Realizability-Preserving DG-IMEX Method for the Two-Moment Model of Fermion Transport \tnoteref{support}

### 2018-02-03 Documents/M1/realizableFermionicM1/sections/DGIMEXscheme.tex
Additions to imex and dg-imex section

We proceed to establish conditions for which $\vect{\cM}_{\bK}^{(ij)}\in\cR$.  
To this end, to simplify notation, we consider the moments
\begin{equation}
  \vect{\cM}_{\bK}^{*}
  =\vect{\cM}_{\bK}-\hat{c}\,\dt\,\big\langle\,\nabla\cdot\vect{\cF}(\vect{\cM}_{h})\,\big\rangle_{\bK},
\end{equation}
where $\vect{\cM}_{\bK}$ is the cell average of $\vect{\cM}_{h}\in\bbV^{k}$ and $\hat{c}\ge0$.  
\begin{lemma}
  Let $\{s_{i}\}_{i=1}^{d}$ be a set of positive constants satisfying $\sum_{i=1}^{d}=1$.  
  If for each $i\in\{1,\ldots,d\}$, 
  \begin{equation}
    \Gamma^{i}\big[\vect{\cM}_{h}\big](\tilde{\vect{x}}^{i})
    :=\f{1}{|K^{i}|}
    \Big[\,\int_{K^{i}}\vect{\cM}_{h}\,dx^{i}-\f{\hat{c}\,\dt}{s_{i}}\big(\,\widehat{\bcF}^{i}(\vect{\cM}_{h})\big|_{\xH^{i}}-\widehat{\bcF}^{i}(\vect{\cM}_{h})\big|_{\xL^{i}}\,\big)\,\Big]\in\cR,
    \label{eq:realizableGamma}
  \end{equation}
  then $\vect{\cM}_{\bK}^{*}\in\cR$.  
\end{lemma}
\begin{proof}
  It is easy to show that $\vect{\cM}_{\bK}^{*}$ can be expressed as the convex combination
  \begin{equation}
    \sum_{i=1}^{d}s_{i}\,\f{1}{|\tilde{\vect{\bK}}^{i}|}\int_{\tilde{\bK}^{i}}\Gamma^{i}\big[\vect{\cM}_{h}\big]\,d\tilde{\vect{x}}^{i}.  
    \label{eq:cellAverageInTermsOfGamma}
  \end{equation}
  The result follows immediately.  
\end{proof}
\begin{rem}
  If a quadrature rule $\tilde{\vect{Q}}^{i}:C^{0}\to\bbR$, with positive weights, and points defined by the set $\tilde{\vect{S}}^{i}$, is used to approximate the integral over $\tilde{\bK}^{i}$ in \eqref{eq:cellAverageInTermsOfGamma}, it is sufficient for \eqref{eq:realizableGamma} to hold in the quadrature points $\tilde{\vect{S}}^{i}\subset\tilde{\bK}^{i}$.  
\end{rem}

>>>>

We proceed to establish conditions for which $\vect{\cM}_{\bK}^{(ij)}\in\cR$.  
To this end, to simplify notation, we consider the moments
\begin{equation}
  \vect{\cM}_{\bK}^{*}
  =\vect{\cM}_{\bK}-\hat{c}\,\dt\,\big\langle\,\nabla\cdot\vect{\cF}(\vect{\cM}_{h})\,\big\rangle_{\bK},
\end{equation}
where $\vect{\cM}_{\bK}$ is the cell average of $\vect{\cM}_{h}\in\bbV^{k}$ and $\hat{c}\ge0$.  
\begin{lemma}
  Let $\{s_{i}\}_{i=1}^{d}$ be a set of positive constants satisfying $\sum_{i=1}^{d}=1$.  
  If for each $i\in\{1,\ldots,d\}$, 
  \begin{equation}
    \vect{\Gamma}^{i}\big[\vect{\cM}_{h}\big](\tilde{\vect{x}}^{i})
    :=\f{1}{|K^{i}|}
    \Big[\,\int_{K^{i}}\vect{\cM}_{h}\,dx^{i}-\f{\hat{c}\,\dt}{s_{i}}\big(\,\widehat{\bcF}^{i}(\vect{\cM}_{h})\big|_{\xH^{i}}-\widehat{\bcF}^{i}(\vect{\cM}_{h})\big|_{\xL^{i}}\,\big)\,\Big]\in\cR,
    \label{eq:realizableGamma}
  \end{equation}
  then $\vect{\cM}_{\bK}^{*}\in\cR$.  
\end{lemma}
\begin{proof}
  It is easy to show that $\vect{\cM}_{\bK}^{*}$ can be expressed as the convex combination
  \begin{equation}
    \sum_{i=1}^{d}s_{i}\,\f{1}{|\tilde{\vect{\bK}}^{i}|}\int_{\tilde{\bK}^{i}}\vect{\Gamma}^{i}\big[\vect{\cM}_{h}\big]\,d\tilde{\vect{x}}^{i}.  
    \label{eq:cellAverageInTermsOfGamma}
  \end{equation}
  The result follows immediately.  
\end{proof}
\begin{rem}
  If a quadrature rule $\tilde{\vect{Q}}^{i}:C^{0}\to\bbR$, with positive weights, and points defined by the set $\tilde{\vect{S}}^{i}$, is used to approximate the integral over $\tilde{\bK}^{i}$ in \eqref{eq:cellAverageInTermsOfGamma}, it is sufficient for \eqref{eq:realizableGamma} to hold in the quadrature points $\tilde{\vect{S}}^{i}\subset\tilde{\bK}^{i}$.  
\end{rem}

Next, we establish conditions for which \eqref{eq:realizableGamma} holds.  
To simplify the notation, we temporarily drop the dimension index $i$, setting $x^{i}=x$, $K^{i}=K$, $\vect{\Gamma}^{i}=\vect{\Gamma}$, etc.  
We let $\hat{Q}_{N}$ denote the $N$-point \emph{Gauss-Lobatto} quadrature rule on the interval $K=(x_{\Low},x_{\Hgh})$, with points
\begin{equation}
  \hat{S}=\left\{x_{\Low}=\hat{x}_{1},\cdots,\hat{x}_{N}=x_{\Hgh}\right\}, 
  \label{eq:quadraturePointsGL}
\end{equation}
and weights $\hat{w}_{q} \in (0,1]$, normalized so that $\sum_{q=1}^{N} \hat{w}_{q} = 1$.  
(The hat is used to denote the Gauss-Lobatto rule, which includes the endpoints of the interval $K$.)
This quadrature integrates polynomials in $x \in \bbR$ with degree $\le2N-3$ exactly.

### 2018-02-03 Documents/M1/realizableFermionicM1/sections/imex.tex
Additions to imex and dg-imex section

The semi-discretization of the moment equations with the DG method given in Eq.~\eqref{eq:semidiscreteDG} results in a system of ordinary differential equations (ODEs) in each element of the form
\begin{equation}
  \dot{\vect{u}}
  =\vect{\cT}(\vect{u})+\f{1}{\tau}\,\vect{\cQ}(\vect{u}),
  \label{eq:ode}
\end{equation}
where $\vect{u}$ are the degrees of freedom evolved with the DG method; i.e., for a test space spanned by $\{\phi_{i}(\vect{x})\}_{i=1}^{N}\in\bbV^{k}$, we let
\begin{equation}
  \vect{u}=\f{1}{|\bK|}\Big(\,\int_{\bK}\vect{\cM}_{h}\,\phi_{1}\,d\vect{x},\int_{\bK}\vect{\cM}_{h}\,\phi_{2}\,d\vect{x},\ldots,\int_{\bK}\vect{\cM}_{h}\,\phi_{N}\,d\vect{x}\,\Big)^{T}.
\end{equation}
The transport term $\vect{\cT}$ is due to the second and third term in the left-hand side of Eq.~\eqref{eq:semidiscreteDG}, while the collision term $\vect{\cQ}$ is due to the right-hand side of Eq.~\eqref{eq:semidiscreteDG}.

In the application of interest to us, the collision term is stiff ($\tau\ll1$) and must be treated with implicit methods, while we can resolve the time scales induced by the transport term, which we will treat with explicit methods; i.e., we will use IMEX methods \cite{pareschiRusso_2005}.  
Until recently, high-order (second or higher order temporal accuracy) positivity-preserving IMEX methods with time step restrictions solely due to the transport operator were not known.  
Chertock et al. \cite{chertock_etal_2015} presented second order accurate IMEX schemes with a correction step.  
The correction step includes the transport operator, and positivity is subject to a time step restriction $\dt\propto1/\sqrt{\tau}$.  
More recently, Hu et al. \cite{hu_etal_2017}, presented new IMEX schemes for a class of BGK-type collision operators with a correction step that does not include the transport operator.  
In this case, positivity is only subject to time step restrictions stemming from the transport operator.  
These IMEX schemes take the following form \cite{hu_etal_2017}
\begin{align}
  \vect{u}^{(i)}
  &=\vect{u}^{n}
  +\dt\sum_{i=1}^{j-1}\tilde{\alpha}_{ij}\,\vect{\cT}(\vect{u}^{(j)})
  +\dt\sum_{i=1}^{j}\alpha_{ij}\,\f{1}{\tau}\,\vect{\cQ}(\vect{u}^{(j)}),
  \quad i=1,\ldots,s, \label{imexStages} \\
  \tilde{\vect{u}}^{n+1}
  &=\vect{u}^{n}
  +\dt\sum_{i=1}^{s}\tilde{w}_{i}\,\vect{\cT}(\vect{u}^{(i)})
  +\dt\sum_{i=1}^{s}w_{i}\,\f{1}{\tau}\,\vect{\cQ}(\vect{u}^{(i)}), \label{imexIntermediate} \\
  \vect{u}^{n+1}
  &=\tilde{\vect{u}}^{n+1}-\alpha\,\dt^{2}\,\f{1}{\tau^{2}}\,\vect{\cQ}'(\vect{u}^{*})\,\vect{\cQ}(\vect{u}^{n+1}), \label{eq:imexCorrection}
\end{align}
To prove the positivity-preserving property of the IMEX scheme, Hu et al. \cite{hu_etal_2017} rewrite the stage values in Eq.~\eqref{imexStages} as
\begin{equation}
  \vect{u}^{(i)}
  =c_{i0}\,\vect{u}^{n}+\sum_{j=1}^{i-1}c_{ij}\Big[\,\vect{u}^{(j)}+\hat{c}_{ij}\,\dt\,\vect{\cT}(\vect{u}^{(j)})\,\Big]
  +a_{ii}\,\dt\,\f{1}{\tau}\,\vect{Q}(\vect{u}^{(i)}),
  \label{eq:imexStagesRewrite}
\end{equation}
where $c_{i0}+\sum_{j=1}^{i-1}c_{ij}=1$.

For globally stiffly accurate (GSA) IMEX schemes $\tilde{\vect{u}}^{n+1}=\vect{u}^{(s)}$

>>>>

In this section we summarize the IMEX scheme used for the realizability-preserving DG-IMEX method for the two-moment model developed in Section~\ref{sec:realizableDGIMEX}.  
The semi-discretization of the moment equations with the DG method given in Eq.~\eqref{eq:semidiscreteDG} results in a system of ordinary differential equations (ODEs) in each element of the form
\begin{equation}
  \dot{\vect{u}}
  =\vect{\cT}(\vect{u})+\f{1}{\tau}\,\vect{\cQ}(\vect{u}),
  \label{eq:ode}
\end{equation}
where $\vect{u}$ are the degrees of freedom evolved with the DG method; i.e., for a test space spanned by $\{\phi_{i}(\vect{x})\}_{i=1}^{N}\in\bbV^{k}$, we let
\begin{equation}
  \vect{u}=\f{1}{|\bK|}\Big(\,\int_{\bK}\vect{\cM}_{h}\,\phi_{1}\,d\vect{x},\int_{\bK}\vect{\cM}_{h}\,\phi_{2}\,d\vect{x},\ldots,\int_{\bK}\vect{\cM}_{h}\,\phi_{N}\,d\vect{x}\,\Big)^{T}.
\end{equation}
The transport term $\vect{\cT}$ is due to the second and third term in the left-hand side of Eq.~\eqref{eq:semidiscreteDG}, while the collision term $\vect{\cQ}$ is due to the right-hand side of Eq.~\eqref{eq:semidiscreteDG}.

In the application of interest to us, the collision term is stiff ($\tau\ll1$) and must be treated with implicit methods, while we can resolve the time scales induced by the transport term, which we will treat with explicit methods; i.e., we will use IMEX methods \cite{pareschiRusso_2005}.  
Until recently, high-order (second or higher order temporal accuracy) positivity-preserving IMEX methods with time step restrictions solely due to the transport operator were not known.  
Chertock et al. \cite{chertock_etal_2015} presented second order accurate IMEX schemes with a correction step.  
The correction step includes the transport operator, and positivity is subject to a time step restriction $\dt\propto1/\sqrt{\tau}$.  
More recently, Hu et al. \cite{hu_etal_2017}, presented new IMEX schemes for a class of BGK-type collision operators with a correction step that does not include the transport operator.  
In this case, positivity is only subject to time step restrictions stemming from the transport operator.  
These $s$-stage IMEX schemes take the following form \cite{hu_etal_2017}
\begin{align}
  \vect{u}^{(i)}
  &=\vect{u}^{n}
  +\dt\sum_{i=1}^{j-1}\tilde{\alpha}_{ij}\,\vect{\cT}(\vect{u}^{(j)})
  +\dt\sum_{i=1}^{j}\alpha_{ij}\,\f{1}{\tau}\,\vect{\cQ}(\vect{u}^{(j)}),
  \quad i=1,\ldots,s, \label{imexStages} \\
  \tilde{\vect{u}}^{n+1}
  &=\vect{u}^{n}
  +\dt\sum_{i=1}^{s}\tilde{w}_{i}\,\vect{\cT}(\vect{u}^{(i)})
  +\dt\sum_{i=1}^{s}w_{i}\,\f{1}{\tau}\,\vect{\cQ}(\vect{u}^{(i)}), \label{imexIntermediate} \\
  \vect{u}^{n+1}
  &=\tilde{\vect{u}}^{n+1}-\alpha\,\dt^{2}\,\f{1}{\tau^{2}}\,\vect{\cQ}'(\vect{u}^{*})\,\vect{\cQ}(\vect{u}^{n+1}), \label{eq:imexCorrection}
\end{align}
where, as in standard IMEX schemes, $(\tilde{a}_{ij})$ and $(a_{ij})$, components of $s\times s$ matrices, and the vectors $(\tilde{w}_{1},\ldots,\tilde{w}_{s})^{T}$ and $(w_{1},\ldots,w_{s})^{T}$ must satisfy certain order conditions \cite{pareschiRusso_2005}.  
The coefficient in the correction step is positive, $\alpha>0$.  
For globally stiffly accurate (GSA) IMEX schemes, $\tilde{w}_{i}=\tilde{a}_{si}$ and $w_{i}=a_{si}$ for $i=\{1,\ldots,s\}$, so that $\tilde{\vect{u}}^{n+1}=\vect{u}^{(s)}$.  
To prove the positivity-preserving property of the IMEX scheme, Hu et al. \cite{hu_etal_2017} rewrite the stage values in Eq.~\eqref{imexStages} as
\begin{equation}
  \vect{u}^{(i)}
  =c_{i0}\,\vect{u}^{n}+\sum_{j=1}^{i-1}c_{ij}\Big[\,\vect{u}^{(j)}+\hat{c}_{ij}\,\dt\,\vect{\cT}(\vect{u}^{(j)})\,\Big]
  +a_{ii}\,\dt\,\f{1}{\tau}\,\vect{Q}(\vect{u}^{(i)}),
  \label{eq:imexStagesRewrite}
\end{equation}
where $c_{i0}$, $c_{ij}$, and $\hat{c}_{ij}=\tilde{c}_{ij}/c_{ij}$ are computed from $\tilde{a}_{ij}$ and $a_{ij}$ (see \cite{hu_etal_2017} for details).  
Importantly, $c_{i0}+\sum_{j=1}^{i-1}c_{ij}=1$.  
The positivity-preserving property follows from requiring $a_{ii}>0$ and $c_{i0}\ge0$, for $i=1,\ldots,s$, and $c_{ij},\tilde{c}_{ij}\ge0$, for $i=2,\ldots,s$, $j=1,\ldots,s-1$.
